Intended Table of Contents Thesis

### 2013-02-08 Chapters/Chapter2.tex
K-means graph
Removed linebreaks

\label{Chapter2} % For referencing the chapter elsewhere, use \ref{Chapter1}

@@

\section{Temporal Segmentation}

This section will give an introduction and in-depth analysis of temporal
segmentation.

\subsection{Aims of segmentation}
When processing and analyzing time series of data, e.g. motion measurements,
stock market fluctuations or natural language, first a low-level division
between the discriminative parts of the stream must be made. One can view this
as splitting the series into the \emph{atoms}, which are the building blocks
of the total stream. These building blocks will be the aggregation of
non-overlapping, internally homogeneous segments \cite{himberg2001time}. This
means that the data points inside a segment should have some resemblance
relation to each other and their difference lies between some boundary. The
process of segmenting can be viewed as a subproblem to context analysis of
time series. Temporal segmentation is closely related to temporal clustering,
although it is a stricter, and simpler, process. Whereby clustering only
restricts the data points on their distance relation (as used in a Voronoi
diagram), within a segment the data points must also be contiguous.

@@

\subsection{Formal definition}
Formally, temporal segmentation is dividing a time series $s$, which consists
of $N$ samples $\mathbf{x}(1),\mathbf{x}(2),\dots,\mathbf{x}(N)$ from
$\mathbf{R}^d$. Individual \emph{segments} are referenced by $s(a,b)$,
consisting of the consecutive samples
$\mathbf{x}(a),\mathbf{x}(a+1),\dots,\mathbf{x}(b)$, $a \le b$. Let $s_1 =
s(a,b)$ and $s_2 = s(b+1,c)$ be two segments, then their concatenation is
$s_1s_2 = s(a,c)$. A segmentation $S$ of $s$ consists of a sequence of $k$
non-empty segments $s_1s_2 \dots s_k = s$.
This notation is adopted from \cite{himberg2001time}.

@@

Many fields of research have been active in the unsupervised segmentation of
data. Many authors rely on a form of Principal Component Analysis (PCA), as
used a.o. in \cite{barbivc2004segmenting}. Often PCA is used to reduce the
dimensionality of the data being processed [REFERENCE] by only using the top
$r$ dimensions to describe the data set. It is observed that data series of
simple motions have a lower dimensionality then complexer motions. When a
simple (repetitive) motion is about to end and fluently transforms in a new
motion, there will be a window of time in which a high dimensionality will be
present, due to the new motion. After this period of transition, the
dimensionality will decrease, since only the new simple motion is present in
the window of time. The first algorithm of \cite{barbivc2004segmenting} is
based on this principle.

Given a set of data points, a lower dimensional hyperplane can by constructed
to which the data points can be projected. This projection on a lower
dimension introduces a error to the original position. When the error is
fixed, less dimensions are needed for simple motions in which movements of
body parts are highly correlated. For segments in which the data points are
lesser correlated, e.g. because of transition state, a higher degree of
dimensions of the hyperplane is needed to represent the data with equal error
degree. When the dimensionality is reduced from $d$ to $r \le d$, the ratio of
error
$E_r$ can be calculated as

\begin{equation}
	E_r = \frac{\sum_{j=1}^{r} \sigma_j^2}{\sum_{j=1}^{d} \sigma_j^2}
\end{equation}

where $\sigma_j$ are the singular values as a result from Singular Value
Decomposition (SVD), which is closely related to PCA \cite{shlens2005tutorial}.

In \cite{barbivc2004segmenting} the stream of frames is analyzed on cut-frames
to find transitions between action. First, for a number of $k$ frames (e.g.
the
equivalent of 2.5 seconds) the required dimensionality $r$ to keep the error
$E_r$
below some threshold $\tau$ is calculated. This will yield in an error $e_i$
for the first
$i$ frames. When more frames are added to the window, the error will increase
with a low constant when it is still in the same activity, due to noise in the
activity. When a new activity
starts, e.g. at frame $j$, the error at frame $j$ will start increasing
faster. This can be expressed by the derivative of the error rate $d_i = e_i -
e_{i-l}$, where $l$ is a constant to remove noise. From this derivative the
mean and standard deviation can be calculated, for each point. When a
derivative $d_j$ rises more than a factor $k_\sigma = 3$ standard deviations
from the mean, a transition point is encountered. The previous frames are then
cut from the sequence (as a segment) and the algorithm starts over.

@@

A second approach in \cite{barbivc2004segmenting} uses the probabilistic
variant of PCA (PPCA) to model the data set as a Gaussian distribution instead
of ignoring the frames which do not fit in the subspace. Over windows of
frames the mean and variance are calculated. In a forward manner the
Mahalanobis distance of a new window of frames is calculated, which represents
the likelihood of the new window belonging to the same segment as the original
widow. When the distance decreases, the likelihood increases which happens
when the motions in the becomes more homogeneous. When a peak in the distance
is reached, the new window of frames indicates a heterogeneous collection of
motions in the window and thus a low likelihood of membership and a indication
of a transition. In order to distinct activities and subactivities (which
require a subset of motions is a distinct activity) the algorithm is also
processed backward over the data series.

@@

The third algorithm in \cite{barbivc2004segmenting} is based on the
observation that data points (frames) tend to form clusters in the space.
These clusters are represented by $k$ Gaussian distributions for which each
the Expectation-Maximization (EM) algorithm estimates the mean $m_j$,
covariance matrix $\sum_{j}$ and prior $\pi_j$. With all the Gaussian
distributions estimated, the data points are assigned to the cluster with the
highest membership likelihood. When two consecutive frames $x_i$ and $x_{i+1}$
belong to different clusters, a transition of activities is recognized. Note
that this algorithm succeeds in segmenting the data and also labels the
similar simple activities.

A drawback in this system, and many others which implement a variant of the
$k$-means algorithm, is that the number of clusters $k$ need to be
predetermined. To cope with this, often the algorithm is performed multiple
times for different values of $k$. Using some criteria, e.g. the Bayesian
Information Criterion \cite{pelleg2000x} or the Davies-Bouldin Index which
guides $k$-means clustering as used in \cite{krause2003unsupervised}.

@@

In \cite{guenterberg2009automatic} a automatic system is used to segment a
continuous stream of data. By calculating features as activity level from the
standard deviation of a sliding window and applying Adaptive Thresholds, the
data can be segmented in active and rest parts. There is no model
representation.

@@

The workings of the PCA \cite{smith2002tutorial} relies on the concepts of
standard deviation, variance,
covariance, eigenvectors and eigenvalues of matrices. These concepts will be
discussed very briefly \footnote{For a more in-depth discussion we would like
to refer the reader to \cite{jolliffe2005principal}}. The standard deviation
and variance
of a set are measures for the
spread around the mean for a single dimension, or feature. The covariance
between two features (dimensions of the data points) indicates how they are
related; when positive the two
features will increase together; when negative one will decrease when the
other increases and when zero they are unrelated. The covariance matrix gives
all the covariances for all pairs of features. This matrix is symmetrical
about the diagonal and the values on the diagonal are the variances for each
feature. For this matrix the eigenvectors can be computed. Eigenvectors have
the characteristic that when they multiply a matrix, the resulting vector is a
multiple of the Eigenvector. The amount by which it is the multiple is the
Eigenvalue. This is illustrated in formulae \ref{eq:no-eigenvector} and
\ref{eq:eigenvector}. The second formula shows an Eigenvector
$ \left( \begin{smallmatrix}
3 \\ 2 \end{smallmatrix} \right)$ and its
associated Eigenvalue, 4. The vector in the first formula is not an
Eigenvector.

@@

With this ordering there are a few applications possible. The first is to just
make a (comprehensible) representation of the data. The Eigenvectors describe
the cloud of data points and thus are a compressed representation of the data.
When the original data points are to be compressed, it is possible to remove
the least significant features and reconstruct the data points from the
resulting Eigenvectors. This will yield in a lossy compression. *** [figure to
illustrate]. An extension of this application is to determine the number of
features needed to keep the compression within a certain error criterion, as
used in \cite{barbivc2004segmenting}. Sets of data can then be distinguished
by the number of features needed to describe the points.

@@

\subsection{Hidden Markov Models}
Many dynamic systems can be viewed as some sequence of consecutive states
which are observable by the world. These kind of systems are known as Markov
Chains. When the state in which the system is can not be observed with
complete certainty, a system known as a Hidden Markov Model (HMM) can be
constructed. It has been shown that HMMs are effective on in the task of
speech recognition \cite{rabiner1989tutorial}. Because of some resemblance
between natural speech construction and human activities, modifications have
been able to classify human activities \cite{guenterberg2009distributed} and
extensions are implemented which create layers of HMMs to handle different
granularity over time and reduce retraining
\cite{oliver2002layered,perdikis2008recognition}. A HMM is a state-space model
representing a single model for each label which needs to be classified.

Formally, a Hidden Markov Model is characterized by the following elements
\cite{rabiner1989tutorial}:
\begin{enumerate}
	\item A number of $N$ states in the model. These states are a
	representation of the underlying mechanism of the system and can not be
	directly observed from outside (hence the \emph{hidden} model). Individual
	states are referenced by $S = \{S_1, S_2, \dots, S_N \}$ and a single
	state in time $t$ is denoted by $q_t$.
	\item An alphabet of $M$ discrete observable symbols for each state. An
	observed symbol is the only indication of the systems current situation.
	Each individual symbol is referenced by $V = \{ v_1, v_2, \dots, v_m \}$.
	\item A state transition probability distribution $A = \{ a_{ij} \}$
	indication the probability of being in state $j$ at time $t+1$ when the
	state on time $t$ was $i$:
	\begin{eqnarray}
		a_{ij} = P [ q_{t+1} = S_j | q_t = S_i ], & 1 \le i, j \le N
	\end{eqnarray}
	Note that by setting the probability $a_{ij}=0$ for some $i$ and $j$
	indicates that $j$ is not reachable from state $i$.
	\item A symbol observation probability distribution in state $j$, $B = \{
	b_j(k) \}$, where
	\begin{eqnarray}
	b_j(k) = P[ v_k\ \mathrm{at} \ t | q_t = S_j], & 1 \le j \le N \nonumber \\
	& 1 \le k \le M
	\end{eqnarray}
	Note that every symbol can be observed in every state.
	\item The initial state distribution in which the system starts $\pi= \{
	\pi_i \}$ where
	\begin{eqnarray}
	\pi_i = P[q_1 = S_i], & 1 \le i \le N
	\end{eqnarray}
\end{enumerate}
This system is often abbreviated to
\begin{equation}
	\lambda = \{ \pi_i, a_{ij}, b_j(k)\}
\end{equation}
The HMM will act as a generator giving a sequence of $T$ observations, each
being a symbol from $V$:
\begin{equation}
	O = O_1, O_2 \dots O_T
\end{equation}

@@

Generally a HMM can be fully connected, or ergodic model. Other configurations
are also possible, of which the left-right model, or Bakis model, is popular
for temporal applications \cite{rabiner1989tutorial}. This because the
structure of the state probability distribution resembles the construction of
possible states over time. This left-right model can be generalized by setting
\begin{eqnarray}
	a_{ij} = 0, & j > i + \Delta
\end{eqnarray}
for some $\Delta$, indicating the possible \emph{distance} of the jumps over
states.

@@

*** Applications of HHM: \cite{shi2009towards} uses 10 coefficients extracted
with FFT from 6 axis (3 times accelerometer and 3 times gyroscope). Example of
continuous HMM (?). Recognition rates fo 50\%-100\%.

In \cite{oliver2002layered} it is stated that HMMs are robust mechanisms with
respect to changes in temporal segmentation of observations, although they
suffer from a few problems when applied to reason about longer and complexer
sequences over time. They state that, especially with limited training data,
the HMM tend to be overfitted and tend to lack structure and have an excess of
parameters. The overcome this limitation, they introduce a Layered Hidden
Markov Model (LHHM) which they show is more robust to temporal changes in
observations. They make use of the hypothesis from the field of psychology,
that human behavior is hierarchically structured \cite{zacks2001event}. In
\cite{perdikis2008recognition} a layered construction of HMMs is used which
feeds the low-level data from a sliding window to the lower layer. This layer
will recognize simple motions, such as moving the arm in a pre-defined
segment. The results from these lower HMMs are then fed to a higher-level
layer of HMMs which are able to recognize abstract motions, such as adjusting
a monitor and picking up a pen. The results showed that the layered system was
better at recognition then a similar single-layered setup. It was also robust
in the sense that when the environment was changed, only the relatively simple
low-level HMMs needed to be retrained.

@@

\subsection{Segmentation as clustering}
In the previous section the discussed methods all relied on the stream of data
points and tried to find cuts the discriminate between successive different
type of activities. An other approach is to consider the tasks of segmentation
as a type of clustering \cite{zhou2008aligned}. In clustering the objective is
to assign labels, or classes, to all the data points indication a similar type
of activity. A clustering $\mathcal{L}$ is thereby more informative then a
segmentation but is also harder to produce.

@@

Data clustering can be used with two different applications, exploratory or
confirmatory \cite{jain1999data}. In the former, the purpose is to discover
clusters in a point cloud of data points. Here the cluster is defined as a
group of homogeneous data points measured by some coherence measure, often a
distance function with the data points being defined as vectors. The produced
clusters, which together form a representation of the data examined, can be
used in the
latter application to assign new provided data points to a cluster and thereby
classify them.

@@

\subsection{k-means clustering}
[EXPLAIN] divide data set $n$ into $k$ clusters.

Among many unsupervised clustering techniques, $k$-means is successfully
applied to large data sets. It is simple to implement and linear in time
complexity so computationally attractive \cite{jain1999data}. A drawback of
the method is that the results of the algorithm greatly depends on the initial
configuration (the data points which will act as centroids) and the number of
cluster $k$ must be determined beforehand.

Generally, the $k$-means methods will minimize the squared error for a
clustering $\mathcal{L}$ criterion which is defined as the distance from the
data points the centroid for each cluster in $\mathcal{K}$. This is expressed
as optimizing to a local optimum the energy function
\begin{equation} \label{eq:k-means energy}
e^2(\mathcal{K},\mathcal{L}) =
\sum_{j=i}^{K}\sum_{i=1}^{n_j}\|\mathbf{x}_i^{(j)} -
\mathbf{c}_j\|^2
\end{equation}

There are several limitation on the $k$-means method. One of these is that
only spherical shapes of cluster can be generated. One of the extensions is
kernel $k$-means \cite{scholkopf1998nonlinear}, which implicitly projects the
data points to a higher dimension and thereby is able to form irregular shaped
cluster.

>>>>

\label{Chapter2} % For referencing the chapter elsewhere, use~\ref{Chapter1}

@@

\section{Temporal Segmentation}
This section will give an introduction and in-depth analysis of temporal segmentation.

\subsection{Aims of segmentation}
When processing and analyzing time series of data, e.g. motion measurements, stock market fluctuations or natural language, first a low-level division between the discriminative parts of the stream must be made.
One can view this as splitting the series into the \emph{atoms}, which are the building blocks of the total stream.
These building blocks will be the aggregation of non-overlapping, internally homogeneous segments~\cite{himberg2001time}.
This means that the data points inside a segment should have some resemblance relation to each other and their difference lies between some boundary.
The process of segmenting can be viewed as a subproblem to context analysis of time series.
Temporal segmentation is closely related to temporal clustering, although it is a stricter, and simpler, process.
Whereby clustering only restricts the data points on their distance relation (as used in a Voronoi diagram), within a segment the data points must also be contiguous.

@@

\subsection{Formal definition}
Formally, temporal segmentation is dividing a time series $s$, which consists of $N$ samples $\mathbf{x}(1),\mathbf{x}(2),\dots,\mathbf{x}(N)$ from $\mathbf{R}^d$.
Individual \emph{segments} are referenced by $s(a,b)$, consisting of the consecutive samples $\mathbf{x}(a),\mathbf{x}(a+1),\dots,\mathbf{x}(b)$, $a \le b$.
Let $s_1 = s(a,b)$ and $s_2 = s(b+1,c)$ be two segments, then their concatenation is $s_1s_2 = s(a,c)$.
A segmentation $S$ of $s$ consists of a sequence of $k$ non-empty segments $s_1s_2 \dots s_k = s$.
This notation is adopted from~\cite{himberg2001time}.

@@

Many fields of research have been active in the unsupervised segmentation of data.
Many authors rely on a form of Principal Component Analysis (PCA), as used a.o. in~\cite{barbivc2004segmenting}.
Often PCA is used to reduce the dimensionality of the data being processed [REFERENCE] by only using the top $r$ dimensions to describe the data set.
It is observed that data series of simple motions have a lower dimensionality then complexer motions.
When a simple (repetitive) motion is about to end and fluently transforms in a new motion, there will be a window of time in which a high dimensionality will be present, due to the new motion.
After this period of transition, the dimensionality will decrease, since only the new simple motion is present in the window of time.
The first algorithm of~\cite{barbivc2004segmenting} is based on this principle.

Given a set of data points, a lower dimensional hyperplane can by constructed to which the data points can be projected.
This projection on a lower dimension introduces a error to the original position.
When the error is fixed, less dimensions are needed for simple motions in which movements of body parts are highly correlated.
For segments in which the data points are lesser correlated, e.g. because of transition state, a higher degree of dimensions of the hyperplane is needed to represent the data with equal error degree.
When the dimensionality is reduced from $d$ to $r \le d$, the ratio of error $E_r$ can be calculated as
\begin{equation}
	E_r = \frac{\sum_{j=1}^{r} \sigma_j^2}{\sum_{j=1}^{d} \sigma_j^2}
\end{equation}

where $\sigma_j$ are the singular values as a result from Singular Value Decomposition (SVD), which is closely related to PCA~\cite{shlens2005tutorial}.

In~\cite{barbivc2004segmenting} the stream of frames is analyzed on cut-frames to find transitions between action.
First, for a number of $k$ frames (e.g. the equivalent of 2.5 seconds) the required dimensionality $r$ to keep the error $E_r$ below some threshold $\tau$ is calculated.
This will yield in an error $e_i$ for the first $i$ frames.
When more frames are added to the window, the error will increase with a low constant when it is still in the same activity, due to noise in the activity.
When a new activity starts, e.g. at frame $j$, the error at frame $j$ will start increasing faster.
This can be expressed by the derivative of the error rate $d_i = e_i - e_{i-l}$, where $l$ is a constant to remove noise.
From this derivative the mean and standard deviation can be calculated, for each point.
When a derivative $d_j$ rises more than a factor $k_\sigma = 3$ standard deviations from the mean, a transition point is encountered.
The previous frames are then cut from the sequence (as a segment) and the algorithm starts over.

@@

A second approach in~\cite{barbivc2004segmenting} uses the probabilistic variant of PCA (PPCA) to model the data set as a Gaussian distribution instead of ignoring the frames which do not fit in the subspace.
Over windows of frames the mean and variance are calculated.
In a forward manner the Mahalanobis distance of a new window of frames is calculated, which represents the likelihood of the new window belonging to the same segment as the original widow.
When the distance decreases, the likelihood increases which happens when the motions in the becomes more homogeneous.
When a peak in the distance is reached, the new window of frames indicates a heterogeneous collection of motions in the window and thus a low likelihood of membership and a indication of a transition.
In order to distinct activities and subactivities (which require a subset of motions is a distinct activity) the algorithm is also processed backward over the data series.

@@

The third algorithm in~\cite{barbivc2004segmenting} is based on the observation that data points (frames) tend to form clusters in the space.
These clusters are represented by $k$ Gaussian distributions for which each the Expectation-Maximization (EM) algorithm estimates the mean $m_j$, covariance matrix $\sum_{j}$ and prior $\pi_j$.
With all the Gaussian distributions estimated, the data points are assigned to the cluster with the highest membership likelihood.
When two consecutive frames $x_i$ and $x_{i+1}$ belong to different clusters, a transition of activities is recognized.
Note that this algorithm succeeds in segmenting the data and also labels the similar simple activities.

A drawback in this system, and many others which implement a variant of the $k$-means algorithm, is that the number of clusters $k$ need to be predetermined.
To cope with this, often the algorithm is performed multiple times for different values of $k$.
Using some criteria, e.g. the Bayesian Information Criterion~\cite{pelleg2000x} or the Davies-Bouldin Index which guides $k$-means clustering as used in~\cite{krause2003unsupervised}.

@@

In~\cite{guenterberg2009automatic} a automatic system is used to segment a continuous stream of data.
By calculating features as activity level from the standard deviation of a sliding window and applying Adaptive Thresholds, the data can be segmented in active and rest parts.
There is no model representation.

@@

The workings of the PCA~\cite{smith2002tutorial} relies on the concepts of standard deviation, variance, covariance, eigenvectors and eigenvalues of matrices.
These concepts will be discussed very briefly \footnote{For a more in-depth discussion we would like to refer the reader to~\cite{jolliffe2005principal}}.
The standard deviation and variance of a set are measures for the spread around the mean for a single dimension, or feature.
The covariance between two features (dimensions of the data points) indicates how they are related; when positive the two features will increase together; when negative one will decrease when the other increases and when zero they are unrelated.
The covariance matrix gives all the covariances for all pairs of features.
This matrix is symmetrical about the diagonal and the values on the diagonal are the variances for each feature.
For this matrix the eigenvectors can be computed. Eigenvectors have the characteristic that when they multiply a matrix, the resulting vector is a multiple of the Eigenvector.
The amount by which it is the multiple is the Eigenvalue.
This is illustrated in formulae~\ref{eq:no-eigenvector} and \ref{eq:eigenvector}.
The second formula shows an Eigenvector $ \left( \begin{smallmatrix} 3 \\ 2 \end{smallmatrix} \right)$ and its associated Eigenvalue, 4.
The vector in the first formula is not an Eigenvector.

@@

With this ordering there are a few applications possible.
The first is to just make a (comprehensible) representation of the data.
The Eigenvectors describe the cloud of data points and thus are a compressed representation of the data.
When the original data points are to be compressed, it is possible to remove the least significant features and reconstruct the data points from the resulting Eigenvectors.
This will yield in a lossy compression. *** [figure to illustrate].
An extension of this application is to determine the number of features needed to keep the compression within a certain error criterion, as used in~\cite{barbivc2004segmenting}.
Sets of data can then be distinguished by the number of features needed to describe the points.

@@

\subsection{Hidden Markov Models}
Many dynamic systems can be viewed as some sequence of consecutive states which are observable by the world.
These kind of systems are known as Markov Chains.
When the state in which the system is can not be observed with complete certainty, a system known as a Hidden Markov Model (HMM) can be constructed.
It has been shown that HMMs are effective on in the task of speech recognition~\cite{rabiner1989tutorial}.
Because of some resemblance between natural speech construction and human activities, modifications have been able to classify human activities~\cite{guenterberg2009distributed} and extensions are implemented which create layers of HMMs to handle different granularity over time and reduce retraining \cite{oliver2002layered,perdikis2008recognition}.
A HMM is a state-space model representing a single model for each label which needs to be classified.

Formally, a Hidden Markov Model is characterized by the following elements
\cite{rabiner1989tutorial}:
\begin{enumerate}
	\item
		A number of $N$ states in the model.
		These states are a representation of the underlying mechanism of the system and can not be directly observed from outside (hence the \emph{hidden} model).
		Individual states are referenced by $S = \{S_1, S_2, \dots, S_N \}$ and a single state in time $t$ is denoted by $q_t$.
	\item
		An alphabet of $M$ discrete observable symbols for each state.
		An observed symbol is the only indication of the systems current situation.
		Each individual symbol is referenced by $V = \{ v_1, v_2, \dots, v_m \}$.
	\item
		A state transition probability distribution $A = \{ a_{ij} \}$ indication the probability of being in state $j$ at time $t+1$ when the state on time $t$ was $i$:
		\begin{eqnarray}
			a_{ij} = P [ q_{t+1} = S_j | q_t = S_i ], & 1 \le i, j \le N
		\end{eqnarray}
		Note that by setting the probability $a_{ij}=0$ for some $i$ and $j$ indicates that $j$ is not reachable from state $i$.
	\item
		A symbol observation probability distribution in state $j$, $B = \{b_j(k) \}$, where
		\begin{eqnarray}
			b_j(k) = P[ v_k\ \mathrm{at} \ t | q_t = S_j], & 1 \le j \le N \nonumber \\
			& 1 \le k \le M
		\end{eqnarray}
		Note that every symbol can be observed in every state.
	\item
	The initial state distribution in which the system starts $\pi= \{ \pi_i \}$ where
	\begin{eqnarray}
		\pi_i = P[q_1 = S_i], & 1 \le i \le N
	\end{eqnarray}
\end{enumerate}

This system is often abbreviated to
\begin{equation}
	\lambda = \{ \pi_i, a_{ij}, b_j(k)\}
\end{equation}
The HMM will act as a generator giving a sequence of $T$ observations, each being a symbol from $V$:
\begin{equation}
	O = O_1, O_2 \dots O_T
\end{equation}

@@

Generally a HMM can be fully connected, or ergodic model.
Other configurations are also possible, of which the left-right model, or Bakis model, is popular for temporal applications~\cite{rabiner1989tutorial}.
This because the structure of the state probability distribution resembles the construction of possible states over time.
This left-right model can be generalized by setting
\begin{eqnarray}
	a_{ij} = 0, & j > i + \Delta
\end{eqnarray}
for some $\Delta$, indicating the possible \emph{distance} of the jumps over states.

@@

*** Applications of HHM:~\cite{shi2009towards} uses 10 coefficients extracted with FFT from 6 axis (3 times accelerometer and 3 times gyroscope).
Example of continuous HMM (?).
Recognition rates fo 50\%-100\%.

In~\cite{oliver2002layered} it is stated that HMMs are robust mechanisms with respect to changes in temporal segmentation of observations, although they suffer from a few problems when applied to reason about longer and complexer sequences over time.
They state that, especially with limited training data, the HMM tend to be overfitted and tend to lack structure and have an excess of parameters.
The overcome this limitation, they introduce a Layered Hidden Markov Model (LHHM) which they show is more robust to temporal changes in observations.
They make use of the hypothesis from the field of psychology, that human behavior is hierarchically structured~\cite{zacks2001event}.
In \cite{perdikis2008recognition} a layered construction of HMMs is used which feeds the low-level data from a sliding window to the lower layer.
This layer will recognize simple motions, such as moving the arm in a pre-defined segment.
The results from these lower HMMs are then fed to a higher-level layer of HMMs which are able to recognize abstract motions, such as adjusting a monitor and picking up a pen.
The results showed that the layered system was better at recognition then a similar single-layered setup.
It was also robust in the sense that when the environment was changed, only the relatively simple low-level HMMs needed to be retrained.

@@

\subsection{Segmentation as clustering}
In the previous section the discussed methods all relied on the stream of data points and tried to find cuts the discriminate between successive different type of activities.
An other approach is to consider the tasks of segmentation as a type of clustering~\cite{zhou2008aligned}.
In clustering the objective is to assign labels, or classes, to all the data points indication a similar type of activity.
A clustering $\mathcal{L}$ is thereby more informative then a segmentation but is also harder to produce.

@@

Data clustering can be used with two different applications, exploratory or confirmatory~\cite{jain1999data}.
In the former, the purpose is to discover clusters in a point cloud of data points.
Here the cluster is defined as a group of homogeneous data points measured by some coherence measure, often a distance function with the data points being defined as vectors.
The produced clusters, which together form a representation of the data examined, can be used in the latter application to assign new provided data points to a cluster and thereby classify them.

@@

\subsection{k-means clustering}
[EXPLAIN] divide data set $n$ into $k$ clusters. Follows the outline in figure~\ref{fig:k_means}

Among many unsupervised clustering techniques, $k$-means is successfully applied to large data sets.
It is simple to implement and linear in time complexity so computationally attractive~\cite{jain1999data}.
A drawback of the method is that the results of the algorithm greatly depends on the initial configuration (the data points which will act as centroids) and the number of cluster $k$ must be determined beforehand.

\begin{figure}[htbp]
	\centering
		\includegraphics{./Figures/graphs/k_means.png}
		\rule{35em}{0.5pt}
	\caption[K-means]{Outline of the $k$-means algorithm.}
	\label{fig:k_means}
\end{figure}

Generally, the $k$-means methods will minimize the squared error for a clustering $\mathcal{L}$ criterion which is defined as the distance from thedata points the centroid for each cluster in $\mathcal{K}$.
This is expressed as optimizing to a local optimum the energy function
\begin{equation}
	\label{eq:k-means energy}
	e^2(\mathcal{K},\mathcal{L}) =
	\sum_{j=i}^{K}\sum_{i=1}^{n_j}\|\mathbf{x}_i^{(j)} -
	\mathbf{c}_j\|^2
\end{equation}

There are several limitation on the $k$-means method.
One of these is that only spherical shapes of cluster can be generated.
One of the extensions is kernel $k$-means~\cite{scholkopf1998nonlinear}, which implicitly projects the data points to a higher dimension and thereby is able to form irregular shaped cluster.
